Building Markov State Models for Periodically Driven Non-Equilibrium Systems

### 2015-02-10 writedown/period.noneq.msm/submit.jctc.reply/paper.tex
CS works on the revise

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\newcommand{\greenc}[1]{{\color{green} #1}}
% \newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\vect}[1]{#1}
\newcommand{\dd}[1]{\textsf{#1}}
% \newcommand{\fwd}[0]{\textrm{fw}}
% \newcommand{\bwd}[0]{\textrm{bw}}
\newcommand{\fwd}[0]{+}
\newcommand{\bwd}[0]{-}
\newcommand{\period}[0]{T_{\textrm{P}}}
\newcommand{\ml}[0]{\mathcal {L}}
\newcommand{\mo}[0]{\mathcal {O}}
\newcommand{\mbp}[0]{\mathbb {P}}
\newcommand{\mc}[0]{\mathcal {C}}
\newcommand{\dt}[0]{\Delta t}
\newcommand{\id}{\mathrm{Id}}
% \newcommand{\myphi}{\boldsymbol\Phi}
% \newcommand{\mymu}{\boldsymbol\mu}
\newcommand{\myphi}{\Phi}
\newcommand{\mymu}{\mu}
\newcommand{\prob}{\textrm{P}}
\newcommand{\dih}{\textrm{dih}}

@@

Following the process described in Section~\ref{sec:build-msm}, we are able to build a 
3-state MSM for the externally driven alanine dipeptide system,
where the quasi-stationary probability distribution $\mu$, the three core sets, and the forward and backward committors are
estimated as described above, and the MSM transition matrices $\hat P$, $\hat{M}$, and $\hat{T}$ are then evaluated using Eq.~\eqref{eq:msm-tmatrix-00}, \eqref{eq:msm-ht-1}, and \eqref{eq:msm-hm-1}, correspondingly.
Alternatively, the MSM  transition matrix $\hat P$ is calculated directly from the NEMD trajectories using Eq.~\eqref{hatP_prob} (very good agreement with the one computed from the committors).
The leading eigenvalues of $P=\myphi^{\top}(T)$ are compared with those of $\hat P$ and $\hat{M}^{-1}\hat{T}$ in
Table~\ref{tab:tmp1}.
Not surprisingly, the two approaches for MSM building are consistent.
The MSM is able to accurately reproduce
the largest non-trivial eigenvalue, which means a precise reproduction
of the longest non-trivial implied timescale. The accuracy of the second non-trivial
timescale is not as good as the first, but is still acceptable.
The reason
for the lower accuracy may also be that
the corresponding timescale is 26.5~ps (calculated by $-T/\log(\lambda_2)$),
which is NOT significantly longer than the temporal resolution given by the period $T=10$~ps of the external driving force.
The difference between the eigenvalues of $\hat P$ and $\hat{M}^{-1}\hat T$
  can be taken as an indication for the non-Markovianity of the milestone process $\hat X_m$. This
  non-Markovianity seems to have stronger influence on the second non-trivial timescale than on the first; this may be caused by shorter decorrelation timescales due to weaker metastability of the core sets involved.
Numerically the two MSM transition matrices are:
\[
\hat{M}^{-1}\hat{T}=\left(\begin{array}{ccc}
   0.860 &   0.133 &   0.008 \\
    0.192  &  0.775  &  0.033\\
    0.019   & 0.066 &   0.915
\end{array}
\right),\qquad \hat{P}=\left(\begin{array}{ccc}
   0.882 &   0.110 &   0.008 \\
    0.158  &  0.815  &  0.028\\
    0.023   & 0.055 &   0.922
\end{array}
\right),
\]
from which we see that  the left-handed alpha-helix conformations of the peptide exhibits the strongest metastability.

@@

A final remark concerning the utilization of the proposed methodology may be in order. 
Based on an MSM for given periodic forcing, optimal control problems may come into play:
Using available methods, such as non-equilibrium
linear response theory~\cite{wang2013linear} or computational alchemy for MSMs ~\cite{schutte2014markov}, one can construct the MSM for slightly changed parameters (e.g., period and amplitude) of the external forcing by appropriate reweighting of the MSM for the given forcing. This, in principle, allows for  answering questions such as the following:
for which parameters of the external forcing does one achieve maximal population of  the left-handed $\alpha$-helix?
Such questions, however, will
be addressed the future studies.

>>>>

\newcommand{\redrevision}[1]{\color{red} #1}
\newcommand{\redc}[1]{{\color{red} #1}}
\newcommand{\bluec}[1]{{\color{blue} #1}}
\newcommand{\greenc}[1]{{\color{green} #1}}
% \newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\vect}[1]{#1}
\newcommand{\dd}[1]{\textsf{#1}}
% \newcommand{\fwd}[0]{\textrm{fw}}
% \newcommand{\bwd}[0]{\textrm{bw}}
\newcommand{\fwd}[0]{+}
\newcommand{\bwd}[0]{-}
\newcommand{\period}[0]{T_{\textrm{P}}}
\newcommand{\ml}[0]{\mathcal {L}}
\newcommand{\mo}[0]{\mathcal {O}}
\newcommand{\mbp}[0]{\mathbb {P}}
\newcommand{\mc}[0]{\mathcal {C}}
\newcommand{\dt}[0]{\Delta t}
\newcommand{\id}{\mathrm{Id}}
% \newcommand{\myphi}{\boldsymbol\Phi}
% \newcommand{\mymu}{\boldsymbol\mu}
\newcommand{\myphi}{\Phi}
\newcommand{\mymu}{\mu}
\newcommand{\prob}{\textrm{P}}
\newcommand{\dih}{\textrm{dih}}

@@

Following the process described in Section~\ref{sec:build-msm}, we are able to build a 
3-state MSM for the externally driven alanine dipeptide system,
where the quasi-stationary probability distribution $\mu$, the three core sets, and the forward and backward committors are
estimated as described above, and the MSM transition matrices $\hat P$, $\hat{M}$, and $\hat{T}$ are then evaluated using Eq.~\eqref{eq:msm-tmatrix-00}, \eqref{eq:msm-ht-1}, and \eqref{eq:msm-hm-1}, correspondingly.
Alternatively, the MSM  transition matrix $\hat P$ is calculated directly from the NEMD trajectories using Eq.~\eqref{hatP_prob} (very good agreement with the one computed from the committors).
The leading eigenvalues of $P=\myphi^{\top}(T)$ are compared with those of $\hat P$ and $\hat{M}^{-1}\hat{T}$ in
Table~\ref{tab:tmp1}.
Not surprisingly, the two approaches for MSM building are consistent.
The MSM is able to accurately reproduce
the largest non-trivial eigenvalue, which means a precise reproduction
of the longest non-trivial implied timescale. The accuracy of the second non-trivial
timescale is not as good as the first, but is still acceptable.
%The  lower accuracy is due to the fact that
%the corresponding timescale of 26.5~ps (calculated by $-T/\log(\lambda_2)$) is \emph{not} significantly longer than the temporal resolution given by the period $T=10$~ps of the external driving force:
%%%CS
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{hitting.eps}
  \caption{Schematic plot of comparison between a continuous and time-discretized trajectories in hitting a core set. }
  \label{fig:hitting}
\end{figure}
The lower accuracy of the second non-trivial timescale stems from the
time-disretization underlying our approach. The associated problem is graphically
illustrated in Fig.~\ref{fig:hitting}.  The milestone process is defined
to change state when the trajectory hits a core set that is not the
one it came from.  In Fig.~\ref{fig:hitting}, the milestone process based on the time-continuous trajectory changes to state ``2'' at the
first hit of core set $C_2$. In comparison, this first hitting
event would not be correctly recognized using the time-discretized trajectory; instead, in the time-discretized case 
the milestone process would change to state ``2'' only at the second
hit of core set $C_2$. This respective error in detecting the correct first hitting time will \emph{not} be significant if it is much longer than the discretization time-step.
However, it will become relevant if both are comparable. In our case, the temporal discretization is given by the period $T=10$~ps, which is not well separated from the second non-trivial time-scale
that is 26.5~ps (calculated by $-T/\log(\lambda_2)$). Therefore, we observe an unavoidable deviation in the second timescale.
In applications to more complex and realistic problems (especially larger systems), in particular for
systems exhibiting timescale separation between the external driving period and the interested implied
timescales, good accuracy is expected. 
%%%

The difference between the eigenvalues of $\hat P$ and $\hat{M}^{-1}\hat T$
  can be taken as an indication for the non-Markovianity of the milestone process $\hat X_m$. This
  non-Markovianity seems to have stronger influence on the second non-trivial timescale than on the first; this may be caused by shorter decorrelation timescales due to weaker metastability of the core sets involved.
Numerically the two MSM transition matrices are:
\[
\hat{M}^{-1}\hat{T}=\left(\begin{array}{ccc}
   0.860 &   0.133 &   0.008 \\
    0.192  &  0.775  &  0.033\\
    0.019   & 0.066 &   0.915
\end{array}
\right),\qquad \hat{P}=\left(\begin{array}{ccc}
   0.882 &   0.110 &   0.008 \\
    0.158  &  0.815  &  0.028\\
    0.023   & 0.055 &   0.922
\end{array}
\right),
\]
from which we see that  the left-handed alpha-helix conformations of the peptide exhibits the strongest metastability.

@@

A final remark concerning the utilization of the proposed methodology may be in order: 
%%%CS
The approach presented herein allows for MSM building for periodic forcings only; an extension
to non-periodic driving forces is not straightforward, because in
general Floquet theory is not applicable to non-periodic driving
forcings. However 
%%%
based on an MSM for given periodic forcing, optimal control problems may come into play:
Using available methods, such as non-equilibrium
linear response theory~\cite{wang2013linear} or computational alchemy for MSMs ~\cite{schutte2014markov}, one can construct the MSM for slightly changed parameters (e.g., period and amplitude) of the external forcing by appropriate reweighting of the MSM for the given forcing. This, in principle, allows for  answering questions such as the following:
for which parameters of the external forcing does one achieve maximal population of  the left-handed $\alpha$-helix?
Such questions, however, will
be addressed the future studies.

### 2015-02-10 writedown/period.noneq.msm/submit.jctc.reply/reply.tex
backup
CS works on the revise

On behalf of the co-author, I would like to resubmit our article:
\textit{``Building Markov State Models for
  Periodically Driven Non-Equilibrium Systems''}
by H.~Wang and C.~Sch\"utte to the
\textit{Journal of Chemical Theory and Computation}.
We gratefuly acknowledge the reviewers for the positive reports, and for
their insight and suggestions, which helped us improving the quality
of the manuscript.  In response to their comments, the revisions are
reported below in details.  To help the referees spot the changes in
the manuscript, we have submitted a separate file (highlight.pdf) with
all changes highlighted in red.

@@

The current theoretical framework firstly reduces to an
time-homogeneous Markov process by the Floquet theory.  The extension
to the non-periodic driving forces is not straightforward, because in
general the Floquet theory is not applicable to for arbitrary driving
force. Nevertheless, the question raised by the referee is very
important, and we intend to address in future research.
\\

@@

The lower accuracy of the second non-trivial time scale stems from the
time-disretization problem in core set hitting, which is graphically
illustrated in Fig.~\ref{fig:tmp1}.  The milestone process is defined
to change state when the trajectory hits a core set that is not the
one it came from.  In Fig.~\ref{fig:tmp1}, if the trajectory is
continuum, then the milestone process changes to state ``2'' at the
first hitting at core set $C_2$. In comparison, the first hitting
event would not be recognized if the trajectory is discretized, and
the milestone process changes to state ``2'' only at the second
hitting at core set $C_2$.\\

This time-discretizations error would not be significant if the
interested time-scale is much longer than the discretized time-step.
However, it would be seen if the they are comparable. In our case, the minimum discretization
time-step is the period $T=10$~ps, which is not well separated from the second non-trivial time-scale
that is 26.5~ps.\\

\begin{table}
  \centering
  \begin{tabular*}{0.9\textwidth}{@{\extracolsep{\fill}}l cccccc}\hline\hline
    & NEMD & MSM & MSM &MSM &MSM &MSM \\
    Time Discr. [ps] & 0.5 &  $10$ &  $20$ & $40$ & $80$ & $120$\\
    $t_1$ [ps] & 100.2 & 104.8 & 108.8 & 115.5 & 127.6 & 142.4 \\\hline\hline
  \end{tabular*}
  \caption{The leading non-trivial time-scale computed by NEMD and MSM at different time-discretizations.
  The smallest time-discretization of MSM is constrained by the period of external driving $T$ that is 10~ps.}
  \label{tab:tmp1}
\end{table}
In order to further illustrate the artifact of time-discretizations,
we discretized the trajectory by increasing time-steps, and compare
the leading time-scale of the derived MSM with the NEMD, see Tab.~\ref{tab:tmp1}.
It can be easily concluded that as the time-discretizations increases and approaches
the leading non-trivial time-scale $t_1 = 100.2$~ps, the accuracy of MSM
in computing $t_1$ decreases. When the time-discretizations increases to $4T = 40$~ps, the relative
error is over 15\%.
\\

In the applications of the more complex and realistic problems, if the
system has time-scale separation between the external driving period and the interested implied
time-scales, then good accuracy is expected. Generally speaking, in
more complex molecular system, the interested time-scales are longer
than the alanine dipeptide toy system. Therefore, we would expect 
higher accuracy in more complex systems if the period of external field is 10~ps.
\\

@@

The discretizations time-step in the milestone MSM approach should be
distinguished from the lag-time in the space full partitioning MSM
approach. Larger time-discretizations in milestone MSM generally gives
worse prediction of implied time-scales, whereas larger lag-time in
space full partitioning MSM leads to better accuracy as long as the
lag-time is smaller than the interested time-scale. The
Chapman-Komolgorov test basically checks the consistency of full
partitioning MSM with different lag-times, and it does not apply to
the milestone MSM approach.

>>>>

On behalf of the co-author, I would like to resubmit our article:
\textit{``Building Markov State Models for
  Periodically Driven Non-Equilibrium Systems''}
by H.~Wang and C.~Sch\"utte to the
\textit{Journal of Chemical Theory and Computation}.
We gratefuly acknowledge the reviewers for the positive reports, and for
their insight and suggestions, which helped us improving the quality
of the manuscript.  In response to their comments, the major questions raised by the referees are
reported below in detail.

We revised the manuscript repsectively and reworked the English completely.
To help the referees spot the major changes in
the manuscript, we have submitted a separate file (highlight.pdf) with
all changes highlighted in red.

We hope that the revised manuscript is fit for publication now.

@@

The theoretical framework presented in the manuscript firstly reduces the NEMD process to a
time-homogeneous Markov process by utilizing Floquet theory.  An extension
to non-periodic driving forces is not straightforward, because in
general Floquet theory is not applicable to arbitrary driving
force. Nevertheless, apart from the aspects related to Floquet theory our general approach does well extend to more general driving forces so that
the question raised by the referee is very relevant: Applications of the approach presented in the manuscript to approach like the one of K. Schulten is one of the final goals. \\

However, one should be aware that our approach is \emph{the first} application of the MSM idea to NEMD. We think that the combination of the results presented here with the approach of the ideas presented in Sarich et al., Markov State Models for Rare Events in Molecular Dynamics. Entropy, 16 (1). pp. 258-286 (2013)  will allow for addressing the question of the referee in future research.\\

@@

The lower accuracy of the second non-trivial timescale stems from the
time-disretization problem in core set hitting, which is graphically
illustrated in Fig.~\ref{fig:tmp1}.  The milestone process is defined
to change state when the trajectory hits a core set that is not the
one it came from.  In Fig.~\ref{fig:tmp1}, the milestone process based on the time-continuous trajectory changes to state ``2'' at the
first hitting at core set $C_2$. In comparison, the first hitting
event would not be recognized using the time-discretized trajectory; in this case 
the milestone process would change to state ``2'' only at the second
hitting at core set $C_2$. This respective error in detecting the correct hitting time would not be significant if the
first hitting time is much longer than the discretization time-step.
However, it will become relevant whenever they are comparable. In our case, the minimum discretization
time-step is the period $T=10$~ps, which is not well separated from the second non-trivial time-scale
that is 26.5~ps. Therefore, we observe an unavoidable deviation in the second timescale.\\

\begin{table}
  \centering
  \begin{tabular*}{0.9\textwidth}{@{\extracolsep{\fill}}l cccccc}\hline\hline
    & NEMD & MSM & MSM &MSM &MSM &MSM \\
    Time Discr. [ps] & 0.5 &  $10$ &  $20$ & $40$ & $80$ & $120$\\
    $t_1$ [ps] & 100.2 & 104.8 & 108.8 & 115.5 & 127.6 & 142.4 \\\hline\hline
  \end{tabular*}
  \caption{The leading non-trivial time-scale computed by NEMD and MSM at different time-discretizations.
  The smallest time-discretization of MSM is constrained by the period of external driving $T$ that is 10~ps.}
  \label{tab:tmp1}
\end{table}
In order to further illustrate the artifact of time-discretizations,
we discretized the trajectory by increasing time-steps, and compare
the \emph{leading} timescale $t_1$ of the derived MSM with the NEMD, see Tab.~\ref{tab:tmp1}.
It can be easily seen that as the discretization time-step increases and approaches
the leading non-trivial time-scale $t_1 = 100.2$~ps, the accuracy of MSM
in computing $t_1$ decreases. 
\\

In applications to more complex and realistic problems (larger systems), in particular for
systems exhibiting timescale separation between the external driving period and the interested implied
timescales, good accuracy is expected (and in fact guaranteed by general MSM theory). Generally speaking, in
more complex molecular system, the  timescales of interest are longer
than in the alanine dipeptide toy system. 
\\

@@

The discretization time-step in the milestoning MSM approach should be
distinguished from the lag-time in the full partition MSM
approach. Larger discretization time-steps in milestoning MSM generally result in
worse prediction of implied timescales (as discussed above), while larger lag-times in
full partition MSMs lead to better accuracy as long as the
lag-time is smaller than the time-scales of interest. The
Chapman-Komolgorov test basically checks the consistency of full
partition MSM with different lag-times, and thus does not apply to
milestone MSMs. See Schütte et al. (2011), Markov State Models Based on Milestoning, J. Chem. Phys., 134 (20). 204105 (2011) for more details.
